---
title: "Intro to the lifecontingencies R package"
author: "Giorgio Alfredo Spedicato, Ph.D C.Stat ACAS"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  beamer_presentation:
    incremental: no
    keep_tex: yes
    theme: "CambridgeUS"
header-includes:
- \usepackage{actuarialsymbol}
- \usepackage{longtable}
- \usepackage{graphicx}
- \usepackage{mystyles}
bibliography: lifecontingenciesBiblio.bib
vignette: |
  %\VignetteIndexEntry{Intro to the lifecontingencies R package} 
  %\VignetteEngine{knitr::rmarkdown}  
  %\usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
library(knitr)
rm(list=ls())
```

#Intro

- The lifecontingencies package [@spedLifecon] will be introduced.
- As first half 2017 it is the first R [@rSoftware] package merging demographic and financial mathematics function in order to perform actuarial evaluation of life contingent insurances (annuities, life insurances, endowments, etc).
- The applied examples will shown: how to load the R package, how to perform basic financial mathematics and demographic calculations, how to price and reserve financial products.

---

- The final example will show how to mix lifecontingencies and demography [@demographyR] function to assess the mortality development impact on annuities.
- The interested readers are suggested to look to the package's vignettes (also appeared in the Journal of Statistical Sofware) for a broader overview. [@dickson2009actuarial; and @mazzoleni2000appunti] provide and introduction of Actuarial Mathematics theory.
- Also [@rmetrics1] and [@charpentierCAS] discuss the software. 


#Loading the package
- The package is loaded using 

```{r load, echo=TRUE, warning=FALSE, message=FALSE}
library(lifecontingencies) #load the package
```

- It requires a recent version of R (>=3.0) and the markovchain package [@markovchainR]. The development version of the package requires also Rcpp package [@RcppR].

---

# Package's Financial Mathematics Functions
# Interest functions

- interest2Discount, discount2Interest: from interest to discount and reverse;
- interest2Intensity, intensity2Interest: from intensity to interest and reverse;
- convertible2Effective, effective2Convertible: from convertible interest rate to effective one.

---

- $(1+i)=\left(1+{\frac{i^{{(m)}}}{m}}\right)^{{m}}=e^{\delta}$
- $e^{{\delta }}=\left(1-{\frac{d^{{(m)}}}{m}}\right)^{{-m}}=(1-d)^{{-1}}$

---

```{r finmat.1}
#interest and discount rates
interest2Discount(i=0.03)
discount2Interest(interest2Discount(i=0.03))
#convertible and effective interest rates
convertible2Effective(i=0.10,k=4)
```

---


# Annuities and future values

- annuity: present value (PV) of an annuity;
- accumulatedValue: future value of constant cash flows;
- decreasingAnnuity, increasingAnnuity: increasing and decreasing annuities.

---

- $a_{\lcroof{n}}={\frac  {1-\left(1+i\right)^{{-n}}}{i}}$ 
- $s_{\lcroof{n}}=a_{\lcroof{n}}*(1+i)^n=\frac{(1+i)^n-1}{i}$
- $\ddot{a}_{\lcroof{n}}=a_{\lcroof{n}}*(1+i)$


---
```{r finmat.2}
annuity(i=0.05,n=5) #due
annuity(i=0.05,n=5,m=1) #immediate
annuity(i=0.05,n=5,k=12) #due, with 
# fractional payemnts
```

---

- $\frac{\ddot{a}_{\lcroof{n}} - n*v^n}{i}=\left( Ia \right)$ 
- $\frac{n-a_{\lcroof{n}}}{i}=\left( Da \right)$ 
- $\left( Ia_{n} \right) +\left( Da_{n} \right) = \left( n+1 \right)*a_{\lcroof{n}}$ 

---
```{r finmat.3}
irate=0.04; term=10
increasingAnnuity(i=irate,n=term)+decreasingAnnuity(i=irate,
n=term)-(term+1)*annuity(i=irate,n=term)
```

---

# Other functions

- presentValue: PV of possible varying CFs. 
- duration, convexity: calculate duration and convexity of any stream of CFs. 

---

- $PV=\sum_{t \in T} CF_t*\left(1+i_t \right )^{-t}$
- $D={\frac{1}{P(0)}}\sum _{t=\tau }^{T}t{\frac{c_t}{(1+r)^{t}}}$
- $C=\frac{1}{P(1+r)^{2}} \sum _{i=1}^{n}{\frac {t_{i}(t_{i}+1)F_{i}}{\left(1+r\right)^{t_{i}}}}$

---

```{r finma.3}
#bond analysis
irate=0.03
cfs=c(10,10,10,100)
times=1:4
#compute PV, Duration and Convexity
presentValue(cashFlows = cfs, 
timeIds = times, 
interestRates = irate)
duration(cashFlows = cfs, 
timeIds = times, i = irate)
convexity(cashFlows = cfs,
timeIds = times, i = irate)
```


# Intro


- Lifecontingencies offers a wide set of functions for demographic analysis;
- Survival and death probabilities, expected residual lifetimes and other function can be easily modeled with
the R package;
- Creation and manipulation of life table is easy as well.

---

# Package's demographic functions
# Table creation and manipulation

- new lifetable or actuarialtable methods.
- print, plot show methods.
- probs2lifetable function to create tables from probabilities

---

- $\{ l_0, l_1, l_2, \ldots, l_{\omega} \}$
- $L_{x}=\frac{l_{x}+l_{x+1}}{2}$
- $q_{x,t} = \frac{d_{x+t}}{l_{x}}$

---


```{r demo.1}
data("demoIta")
sim92<-new("lifetable",x=demoIta$X, 
           lx=demoIta$SIM92, name='SIM92')
getOmega(sim92)
tail(sim92)
```

---



# Life tables' functions

- dxt, deaths between age x and x + t, tdx
- pxt, survival probability between age x and x + t, tpx
- pxyzt, survival probability for two (or more) lives, tpxy
- qxt, death probability between age x and x + t, tqx
- qxyzt, death probability for two (or more) lives, tqxy
- Txt, number of person-years lived after exact age x, tTx
- mxt, central death rate, tmx
- exn, expected lifetime between age x and age x + n, nex
- exyz, n-year curtate lifetime of the joint-life status

---

- $p_{x,t} = 1 - q_{x,t} = \frac{l_{x+t}}{l_{x}}$
- $e_{x,n}=\sum_{t=1}^{n}p_{x,t}$

---

```{r demo.2}
#two years death rate
qxt(sim92, x=65,2)
#expected residual lifetime between x and x+n
exn(sim92, x=25,n = 40)
```


---

# Simulation

- rLife, sample from the time until death distribution underlying a life table
- rLifexyz, sample from the time until death distribution underlying two or more lives

---
```{r demo.3}
#simulate 1000 samples of residual life time
res_lt<-rLife(n=1000,object = sim92,x=65)
hist(res_lt,xlab="Residual Life Time")
```


#Assessing longevity impact on annuities using lifecontingencies and demography

- This part of the presentation will make use of the demography package to calibrate Lee Carter [@Lee1992] model, $log\left(\mu_{x,t} \right) =a_{x}+b_{x}*k_{t}\rightarrow p_{x,t}=exp^{-\mu_{x,t}}$ ,projecting mortality and implicit life tables.

```{r leecarter01, tidy=FALSE, include=FALSE, results='hide'}
#library(demography)
#italy.demo<-hmd.mx("ITA", username="spedicato_giorgio@yahoo.it", password="mortality")
```


```{r leecarter0, tidy=FALSE, warning=FALSE, message=FALSE}
#load the package and the italian tables
library(demography) 
#italyDemo<-hmd.mx("ITA", username="yourUN", 
#password="yourPW")
load(file="mortalityDatasets.RData") #load the dataset
```

---

- Lee Carter model is calibrated using lca function of demography package.
- Then an arima model is used to project (extrapolate) the underlying $k_t$ over the historical period.


```{r leecarter1, tidy=FALSE, warning=FALSE}
#calibrate lee carter
italy.leecarter<-lca(data=italyDemo,series="total",
                     max.age=103,adjust = "none")
#perform modeling of kt series
kt.model<-auto.arima(italy.leecarter$kt)
#projecting the kt
kt.forecast<-forecast(kt.model,h=100) 
```


---


-The code below generates the matrix of prospective life tables

```{r leecarter2, tidy=FALSE, size='tiny'}
#indexing the kt
kt.full<-ts(union(italy.leecarter$kt, kt.forecast$mean),
            start=1872)  
#getting and defining the life tables matrix
mortalityTable<-exp(italy.leecarter$ax
+italy.leecarter$bx%*%t(kt.full)) 
rownames(mortalityTable)<-seq(from=0, to=103)
colnames(mortalityTable)<-seq(from=1872, 
to=1872+dim(mortalityTable)[2]-1)
```

---


```{r leecarter2plot, tidy=FALSE, echo=FALSE}
plot.ts(kt.full, main="historical and projected KT",xlab="year",
        ylab="kt",col="steelblue")
abline(v=2009,col="darkred",lwd=2.5)
```

---


- now we need a function that returns the one-year death probabilities
given a year of birth (cohort.


```{r leecarter3, tidy=FALSE}
getCohortQx<-function(yearOfBirth)
{
  colIndex<-which(colnames(mortalityTable)
                  ==yearOfBirth) #identify 
  #the column corresponding to the cohort 
  #definex the probabilities from which 
  #the projection is to be taken
  maxLength<-min(nrow(mortalityTable)-1,
                 ncol(mortalityTable)-colIndex)
  qxOut<-numeric(maxLength+1)
  for(i in 0:maxLength)
    qxOut[i+1]<-mortalityTable[i+1,colIndex+i]
  #fix: we add a fictional omega age where 
  #death probability = 1
  qxOut<-c(qxOut,1)
  return(qxOut)
}
```

---

- Now we use such function to obtain prospective life tables and to perform 
actuarial calculations. For example, we can compute the APV of an annuity on a 
workers' retiring at 65 assuming he were born in 1920, in 1950 and in 1980. We will use the 
interest rate of 1.5\% (the one used to compute Italian Social Security annuity factors).
- The first step is to generate the life and actuarial tables

```{r leecarter4, tidy=FALSE, size='scriptsize'}
#generate the life tables
qx1920<-getCohortQx(yearOfBirth = 1920)
lt1920<-probs2lifetable(probs=qx1920,type="qx",
name="Table 1920")
at1920<-new("actuarialtable",x=lt1920@x,
lx=lt1920@lx,interest=0.015)
qx1950<-getCohortQx(yearOfBirth = 1950)
lt1950<-probs2lifetable(probs=qx1950,
type="qx",name="Table 1950")
at1950<-new("actuarialtable",x=lt1950@x,
lx=lt1950@lx,interest=0.015)
qx1980<-getCohortQx(yearOfBirth = 1980)
lt1980<-probs2lifetable(probs=qx1980,
type="qx",name="Table 1980")
at1980<-new("actuarialtable",x=lt1980@x,
lx=lt1980@lx,interest=0.015)
```

---

- Now we can evaluate $\ddot{a}_{65}$ and $\mathring{e}_{65}$ for workers born in 1920, 1950 and 1980 respectively.


```{r leecarter5, tidy=FALSE, echo=TRUE}
cat("Results for 1920 cohort","\n")
c(exn(at1920,x=65),axn(at1920,x=65))
cat("Results for 1950 cohort","\n")
c(exn(at1950,x=65),axn(at1950,x=65))
cat("Results for 1980 cohort","\n")
c(exn(at1980,x=65),axn(at1980,x=65))
```

---


# Intro on Actuarial Mathematics Funcs


- The lifecontingencies package allows to compute all classical life contingent insurances. 
- Stochastic calculation varying expected lifetimes are possible as well. 
- This makes the lifecontingencies package a nice tool to perform actuarial computation at command line on life insurance tasks. 

---


# Creating actuarial tables

- Actuarial tables are stored as S4 object. 
- The $l_x$, $x$, interest rate and a name are required.
- The print method return a classical actuarial table (commutation functions)


---

```{r actmat.1}
data("demoIta")
sim92Act<-new("actuarialtable",x=demoIta$X, 
lx=demoIta$SIM92, name='SIM92')
sif92Act<-new("actuarialtable",x=demoIta$X,
lx=demoIta$SIF92, name='SIF92')
head(sim92Act)
```

---


# Life insurances functions
## Classical life contingent insurances

- Exn, pure endowment: $\Ax{\pureendow{x}{n}}$
- axn, annuity: $\ddot{a}_{x} = \sum_{k=0}^{\omega-x} v^{k} * p_{x,k}=\sum_{k=0}^{\omega-x} \ddot{a}_{\lcroof{K+1}} p_{x,k} q_{x+k,1}$
- Axn, life insurance: $\Ax{\termxn}=\sum_{k=0}^{n-1} v^{k-1} * p_{x,k} * q_{x+k,1}$
- AExn, endowment: $\Ax{\endowxn}=\Ax{\pureendow{x}{n}}+\Ax{\termxn}$

---

```{r actmat.2, echo=TRUE}
100000*Exn(sim92Act,x=25,n=40)
100000*AExn(sim92Act,x=25,n=40)
1000*12*axn(sim92Act,x=65,k=12)
100000*Axn(sim92Act,x=25,n=40)
```
---

## Additional life contingent insurances


- Increasing and decreasing term insurances and annuities 
- $\left(n-1\right) * \Ax{\termxn} = \IA_{\termxn} + \DA_{\termxn}$

---

```{r actmat.3, echo=TRUE}
IAxn(sim92Act,x=40,n=10)+DAxn(sim92Act,x=40,n=10)
(10+1)*Axn(sim92Act,x=40,n=10)
```
---


## Insurances on multiple lifes

- First survival and last survival status, both for insurances and annuities
- $A_{xy}$+$A_{\overline{xy}}$=$A_x$+$A_y$
- $a_{xy}$+$a_{\overline{xy}}$=$a_x$+$a_y$


---
```{r actmat.5, echo=TRUE}
fr_pay=12
1000*fr_pay*axyzn(tablesList = list(sim92Act,sif92Act),
x = c(64,67),status="last",k=fr_pay)
1000*fr_pay*(axn(sim92Act,x=64,k=fr_pay)+axn(sif92Act,
x=67,k=fr_pay)-axyzn(tablesList = list(sim92Act,sif92Act),
x = c(64,67),status="joint",k=fr_pay))
```


---


# Simulation

- It is possible to simulate the PV of insured benefit distributions. 
- The rLifeContingencies function is used for single life benefit insurance. 
- The rLifeContingenciesXyz function is used for multiple lifes benefits. 

---
```{r act.sim, echo=TRUE}
hist(rLifeContingencies(n = 1000,lifecontingency = "Axn",
x = 40,object = sim92Act,getOmega(sim92Act)-40),
main="Life Insurance on 40 PV distribution")
```
---



# Bibliography {.allowframebreaks}
